ScPy: A SuperCollider Extension for Performing Numerical Computation via Embedded Python

### 2016-08-28 doc/paper.tex
editing, writing, and diagram

\usepackage{fancyhdr}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{minted}
\usepackage[style=numeric-comp]{biblatex}
\usepackage{booktabs}
\usepackage{graphicx}

@@

One of the goals of this project is to enable more flexible experimentation with FFT and phase
vocoder based operations. SuperCollider contains a total of 32 built in phase vocoder operations,
which provide the ability to produce many simple effects. There are additionally some extra
operations available as user created extensions. However, many conceivable effects are impossible,
including basic phase vocoder effects such as pitch shifting. Some examples of novel effects can
be found in~\autoref{sec:usage}.

@@

Audio morphing is defined as a gradual change from one sound to the other. While morphing can be
achieved to some extent simply by cross fading two signals, perceptually this sounds simply like
one sound is becoming quieter as the other is becoming louder. Sethares and Bucklew
\cite{sethares2012kernel} propose density and kernel based techniques that aim to provide a more
convincing morph operation where there is no clear separation between the two sounds at any point
during the morphing process. The method allows the use of different kernel functions to provide
different kinds of morphing. These include \textit{Cross-fade}, \textit{Heat equation}, and
\textit{Harmonic integrity}. The morphing technique is implemented using a phase vocoder.

As a future extension to ScPy, these morphing techniques could be implemented in Python.

As Python is an easily extendible general purpose language, it has applications in many scientific
fields.  Glover et al.~\cite{glover2011python} present a discussion of the use of Python for audio
signal processing. They note that while audio processing libraries exist for many languages,
scripting languages such as Python or MATLAB see widespread adoption because of the speed at which
prototypes can be built. General purpose Python libraries such as NumPy, SciPy and Matplotlib
include many essential functions for audio processing, such as file IO, Fourier transforms, signal
processing and interpolation. There are additionally more specific libraries such as Simpl, a
sinusoidal modelling library, and Modal, an onset detection library.

Any of these libraries can easily be used within ScPy simply by installing and importing them.

@@

@@

@@

TODO data flow diagram

@@

Support for more SuperCollider types, such as events or strings, could be added to open up ScPy to
more use cases.

Multi-core processing support would allow for many more sounds to be processed in real-time.

Creating a more generic extension which supports many languages would have a number of advantages.
One option for implementing this would be to define a specification to communicate and share data
with a generic UGen in a running SuperCollider instance.

Although extending SuperCollider was the focus of this project, similar improvements could be made
to the CSound\cite{boulanger2000csound,lazzarini2005extensions}, Chuck\cite{wang2008chuck} or
Faust\cite{orlarey2009faust} environments as well.

\printbibliography{}

>>>>

\usepackage{fancyhdr}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{minted}
\usepackage[style=numeric-comp]{biblatex}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{tikz}

\usetikzlibrary{arrows}

@@

One of the goals of this project is to enable more flexible experimentation with FFT and phase
vocoder based operations. SuperCollider contains a total of 32 built in spectral operations,
which provide the ability to produce many simple effects. There are additionally some extra
operations available as user created extensions. However, many conceivable effects are impossible,
including basic phase vocoder effects such as pitch shifting. Some examples of novel effects can
be found in~\autoref{sec:usage}.

@@

Audio morphing is defined as a gradual change from one sound to the other. While morphing can be
achieved to some extent simply by cross fading two signals, perceptually this sounds simply like
one sound is becoming quieter as the other is becoming louder. Sethares and Bucklew
\cite{sethares2012kernel} propose density and kernel based techniques that aim to provide a more
convincing morph operation where there is no clear separation between the two sounds at any point
during the morphing process. The method allows the use of different kernel functions to provide
different kinds of morphing. These include \textit{Cross-fade}, \textit{Heat equation}, and
\textit{Harmonic integrity}. The morphing technique is implemented using a phase vocoder. These
techniques could be implemented in Python as a future extension to ScPy.

As Python is an easily extendible general purpose language, it has applications in many scientific
fields. Glover et al.~\cite{glover2011python} present a discussion of the use of Python for audio
signal processing. They note that while audio processing libraries exist for many languages,
scripting languages such as Python or MATLAB see widespread adoption because of the speed at which
prototypes can be built. General purpose Python libraries such as NumPy, SciPy and Matplotlib
include many essential functions for audio processing, such as file IO, Fourier transforms, signal
processing and interpolation. There are additionally more specific libraries such as Simpl, a
sinusoidal modelling library, and Modal, an onset detection library. Any of these libraries can
easily be used within ScPy simply by installing and importing them.

@@

Control UGens only provide a single floating point value per cycle so it would seem natural to
convert them to a Python \texttt{float}, but in ScPy they are instead transferred to Python in a
one element \texttt{ndarray}. This is because Python provides no numerical types with interior
mutability. In practice, NumPy accepts a one element \texttt{ndarray} almost anywhere a
\texttt{float} is accepted, so this detail is not an annoyance.

@@

\begin{figure}[ht]
    \caption{A high level overview of data flow and processing.}
    \begin{center}
        \begin{tikzpicture}[
            every matrix/.style={ampersand replacement=\&,column sep=2cm,row sep=2cm},
            sink/.style={draw,thick,rounded corners,fill=green!20,inner sep=.3cm},
            process/.style={draw,thick,rounded corners,fill=blue!20,inner sep=.3cm},
            datastore/.style={draw,thick,inner sep=.3cm},
            to/.style={->,>=stealth',shorten >=1pt,semithick},
            both/.style={<->,>=stealth',shorten >=1pt,semithick},
            every node/.style={align=center}]

            \matrix{\node[sink] (sc) {SuperCollider}; \\
                    \node[datastore] (objects) {buffer}; \& \node[process] (scpy) {ScPy}; \\
                    \& \node[sink] (py) {Python}; \\
            };

            \draw[to] (sc) -| node[near end,right] {next()} (scpy);
            \draw[to] (objects) |- node[near start,left] {data pointer} (py);
            \draw[both] (objects) -- node[midway,left] {copy data} (sc);
            \draw[to] (scpy) -- node[midway,right] {execute()} (py);
            \draw[to] (scpy) -- node[midway,above] {update()} (objects);
        \end{tikzpicture}
    \end{center}
\end{figure}

@@

@@

Support for more SuperCollider types, such as events or strings, could be added to open up ScPy to
more use cases. Adding support for more types should be trivial in the existing code base.

Multi-core processing support would allow for many more sounds to be processed in real-time.
Currently, all processing is done on a single thread, even with many UGens running at once. Using
the supernova server~\cite{blechmann2005supernova} may be helpful, however it has not yet been
tested and ScPy may require some modifications to be usable with supernova.

Creating a more generic extension which supports many languages would have a number of advantages.
Although extending SuperCollider was the focus of this project, similar improvements could be made
to the CSound~\cite{boulanger2000csound,lazzarini2005extensions}, Chuck~\cite{wang2008chuck} or
Faust~\cite{orlarey2009faust} environments as well. It may be desirable to completely separate the
embedded language from the host, in order to allow easier interoperation and a cleaner
implementation.

\newpage\printbibliography{}
